---
title: Analyzing State Legislator Effectiveness, Part One
subtitle: Empirical Bayesian Shrinkage towards a Beta Prior
author: Corban
date: '2019-05-07'
slug: washington-legislator-efficiency
categories:
  - R
tags: []
image:
  caption: ''
  focal_point: ''
---

Washington State's legislative session recently concluded. The Democrats, who for the second year in a row controlled both the House and the Senate, celebrated success with the passage of signature legislation such as a raise in the smoking age, a public option for health insurance, and environmental protections. However, those big-name bills don't tell the whole story. For this analysis, I wanted to dig in to the over 2500 bills introduced this session by members. I'm also working through David Robinson's *Introduction to Empirical Bayes*, which is a fantastic applied reference to my trip down Bayesian Lane this academic quarter. I equated the baseball "batting average" metric used by Robinson throughout the text to a legislative batting average, being the number of bills a member prime sponsored divided by the total number of bills they sponsored. Emperical Bayesian shrinkage was appropriate in this case, as some members introduced relatively bills that had a higher likelihood of success, without giving credence to their overall numerical significance. Let's see how it looks.  

```{r warning=FALSE, error=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(stats4)
library(VGAM)
library(kableExtra)
library(plotly)
library(hrbrthemes)
library(widgetframe)

bill_data <- read_csv(here("static/data/session_data.csv"))
party_data <- read_csv(here("static/data/leg_rosters.csv"))


bill_data <- bill_data %>%
  separate(Bill, c("prefix", "bill_number"), sep = " ") %>% #seperate e.g. HB 1065 into components
  mutate(billpassed = if_else(Status == "Del to Gov" | startsWith(Status, "C"),"passed", "not_passed")) %>% #filter for bills that passed 
  distinct() %>% #removes house of origin duplicates
  left_join(party_data, by = c("Original Sponsor" = "Member Name")) %>% #join by party affiliation
  select(bill_number, Status, Title,`Original Sponsor`, billpassed, Party) %>% 
  filter(`Original Sponsor` != "People of the State of Washington") %>%  #not including initiatives, sorry Tim Eyman
  mutate(Party = as.factor(Party))

sponsor_record <- bill_data %>% 
  count(`Original Sponsor`, billpassed, Party) %>% 
  spread(billpassed, n) %>% 
  mutate(passed = replace_na(passed, 0),
         not_passed = replace_na(not_passed, 0),
         tot_bills = passed + not_passed,
         pct_passed = passed / (tot_bills))

billsgraph <- ggplot(sponsor_record, aes(x = passed, y = not_passed, color = Party, label = `Original Sponsor`))+
  geom_jitter() +
  labs(title = "Overall Bill Passage",
       y = "Bills not passing",
       x = "Bills passing") +
  theme_modern_rc() +
  scale_color_manual(breaks = c("D", "R"),
                        values=c("blue", "red"))
  

g <- ggplotly(billsgraph)
frameWidget(g)
```

Intuition generally holds up here. The minority party, in this case Republicans, generally passes fewer bills, with a higher volume of bills not passing. However, ranking members from most to fewest bills passed doesn't give us the true picture of effectiveness, because the total number of bills by a given member varies accross the board. In order to get a more complete picture, we can use a Bayesian prior of the overall distribution, and *shrink* the ratio of passes to failures. 

Here's a quick histogram of the distribution of proportion of bills passed.

```{r}
#prior estimation
hist(sponsor_record$pct_passed)
```

Based on our distribution, we can develop estimates of alpha and beta based on the technique outlined in *Empirical Bayes*. 

```{r}
# log-likelihood function
ll <- function(alpha, beta) {
x <- sponsor_record$passed
total <- sponsor_record$tot_bills
-sum(VGAM::dbetabinom.ab(x, total, alpha, beta, log = TRUE))
}

# maximum likelihood estimation
m <- mle(ll, start = list(alpha = 1, beta = 10), method = "L-BFGS-B", lower = c(0.0001, .1))
ab <- coef(m) 
alpha0 <- ab[1]
beta0 <- ab[2]

z <- seq(0,1,length=21)
beta_dist <- data.frame(cbind(z, dbeta(z,alpha0,beta0)))
colnames(beta_dist) <- c("x", "value")

bills_eb <- sponsor_record %>%
mutate(eb_estimate = (passed + alpha0) / (tot_bills + alpha0 + beta0),
       delta = pct_passed-eb_estimate) %>% 
  arrange(eb_estimate) %>% 
  mutate(`Original Sponsor` = as.factor(`Original Sponsor`))


g3 <- ggplot(bills_eb, aes(x = pct_passed, y = eb_estimate, label = `Original Sponsor`, label2 = passed, label3 = not_passed, color = tot_bills, shape = Party)) +
  geom_point() +
  scale_color_viridis_c(name = "Total Bills Introduced") +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Overall Bill Passage",
       y = "Bayesian estimate",
       x = "Observed percentage of bills passed") +
  geom_abline(intercept = 0, slope = 1, color = "lightgray", alpha = .5) +
  theme_modern_rc()
  

g1 <- ggplotly(g3)
frameWidget(g1)
```

We can see the empirical Bayes estimates for those with more bills passed are generally closer to the true observed number of bills passed. Due to the partisan composition of Washington's legislature, democrats tended to have more bills passed and thus are less affected by the empirical Bayes shrinkage. 

Thanks again to David Robinson for the fantastic reference material! Below you will find a list of top and bottom five legislators based on the empirical Bayes shrinkage- it turns out that 76(!) bills introduced and not passed with two passes is enough to rank below those with zero overall passes.

```{r}

bills_eb <- bills_eb %>% 
  arrange(desc(eb_estimate))
kable(tail(bills_eb))
kable(head(bills_eb))
```



